---
title: "Biometric Tracker"
output:
  flexdashboard::flex_dashboard: null
  orientation: columns
  vertical_layout: fill
runtime: shiny
resource_files:
- raw_data/*
- data/*
---
  
```{r setup, include=FALSE}
library(gt)
library(xts)
library(zoo)
library(glue)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)
library(ggthemes)
library(dygraphs)
library(lubridate)
library(tidyverse)
library(flexdashboard)
library(futile.logger)

# load code from other files in this package
source("globals.R")

# in this section of the code we will do all our data reading, cleaning, wrangling...basically
# everything except the timeseries forecasting bit so that the rest of the sections simply display the 
# charts based on the data analysis done here. The forecasting is left to its own section later in the code
# because it is based on user input and so it needs to be done redone whenever the input changes.

# setup logging so that traces are shown on the console and in a log file as well
flog.appender(appender.tee(LOG_FILE), name=LOGGER)

# print the parameters used
flog.info(glue("params used:
                START_DATE={START_DATE},
                DATA_DIR={DATA_DIR},
                P1_NAME={P1_NAME},
                P2_NAME={P2_NAME},
                P1_DATA_FPATH={P1_DATA_FPATH},
                P2_DATA_FPATH={P2_DATA_FPATH},
                CAPTION={CAPTION},
                IMPORTANT_DATES_FNAME={IMPORTANT_DATES_FNAME},
                IMPORTANT_DATES_FPATH={IMPORTANT_DATES_FPATH},
                NUDGE_X={NUDGE_X},
                NUDGE_Y={NUDGE_Y},
                P1_TARGET_WEIGHT={P1_TARGET_WEIGHT},
                P1_WEIGHT_CAP={P1_WEIGHT_CAP},
                P1_WEIGHT_FLOOR={P1_WEIGHT_FLOOR},
                P2_TARGET_WEIGHT={P2_TARGET_WEIGHT},
                P2_WEIGHT_CAP={P2_WEIGHT_CAP},
                P2_WEIGHT_FLOOR={P2_WEIGHT_FLOOR}"), name=LOGGER)


# read the raw data for person 1, print basic summary and metadata
df_P1 <- read_csv(P1_DATA_FPATH) %>%
  mutate(name=P1_NAME) %>%
  mutate(Date = ymd(Date)) %>%
  arrange(Date) %>%
  filter(Date >= START_DATE)
flog.info(glue("read data for {P1_NAME} from {P1_DATA_FPATH}, shape of dataframe={nrow(df_P1)}x{ncol(df_P1)}"), name=LOGGER)
flog.info(head(df_P1), name=LOGGER)
flog.info(summary(df_P1), name=LOGGER)

# read the raw data for person 2, ultimately we want to have this dashboard work the same way
# even if there was only person 1 so put the following in an if checl
if(!is.na(P2_NAME)) {
  df_P2 <- read_csv(P2_DATA_FPATH) %>%
    mutate(name=P2_NAME) %>%
    mutate(Date = ymd(Date)) %>%
    arrange(Date) %>%
    filter(Date >= START_DATE)
  flog.info(glue("read data for {P2_NAME} from {P2_DATA_FPATH}, shape of dataframe={nrow(df_P2)}x{ncol(df_P2)}"), name=LOGGER)
  flog.info(head(df_P2), name=LOGGER)  
  flog.info(summary(df_P2), name=LOGGER)  
}

# read the important dates csv file. This is needed because we would like to annotate this journey
# so that we can say oh right there was an increase in weight for these days and it followed a birthday party, for example...
if(!is.na(IMPORTANT_DATES_MOBILE_FNAME)) {
  important_dates <- read_csv(IMPORTANT_DATES_MOBILE_FPATH)
  flog.info(glue("read data for important dates from {IMPORTANT_DATES_FNAME},
                  shape of dataframe={nrow(important_dates)}x{ncol(important_dates)}"), name=LOGGER)
  flog.info(head(important_dates), name=LOGGER)  
  flog.info(summary(important_dates), name=LOGGER)  
}

# combine the dataframes, we want to do a side by side analysis for both people
if(!is.na(df_P2)) {
  df <- bind_rows(df_P1, df_P2)
  (df) %>%
    sample_n(5)
  flog.info(glue("combined data for {P1_NAME} and {P2_NAME}, shape of data is {nrow(df)}x{ncol(df)}"), name=LOGGER)
} else {
  flog.info(glue("only {P1_NAME} specified, only analyzing data for one person"), name=LOGGER)
  df <- df_P1
}

# get the data in tidy format i.e. Each variable must have its own column.
# Each observation must have its own row.
# Each value must have its own cell.
# see https://r4ds.had.co.nz/tidy-data.html
flog.info("converting the data to tidy format", name=LOGGER)
df_tidy <- df %>%
  gather(metric, value, -Date, -name) %>%
  mutate(value=as.numeric(value))
df_tidy %>%
  sample_n(5)
flog.info(glue("shape of the tidy dataframe is {nrow(df_tidy)}x{ncol(df_tidy)}"), name=LOGGER)

# determine the per day weight loss dataframe by
# calculating loss as weight - the one previous value of weight
# this is done by first grouping the dataframe by name since it has
# data for two people and then arranging by date while maintaining
# the grouping (NOTE: .by_group=TRUE)
df_wt_loss <- df_tidy %>%
  filter(metric=="Weight") %>%
  select(name, Date, value) %>%
  group_by(name) %>%
  arrange(Date, .by_group=TRUE) %>%
  mutate(loss_per_day = -1*(value-lag(value, 1)))  %>%
  mutate(loss_per_day_7_day_ma=rollapply(loss_per_day, 7, mean,align='right',fill=NA))

# is the curse of the weekend real? Assign the day to each date so that we can determine
# if say the weight loss eventually after the weekend was very less or maybe not even there...
df_wt_loss <- df_wt_loss %>%
  mutate(day = weekdays(as.Date(Date)))

# determine how much of theweight loss target has been achieved, this is done by finding the starting
# weight (configured), target weight (configured) and seeing how far each person has reached based on
# what their current weight is. This percentage is used to display a gauge (like the needle of a speedometer)
p1_starting_weight <- df_tidy %>% filter(name==P1_NAME & metric=="Weight") %>% head(1) %>% pull(value)
p1_latest_weight <- df_tidy %>% filter(name==P1_NAME & metric=="Weight") %>% tail(1) %>% pull(value)
# weight loss would be negative when calculated so multiply by -1
p1_wt_lost_as_pct <- -1*100*((p1_latest_weight-p1_starting_weight)/p1_starting_weight)

p2_starting_weight <- df_tidy %>% filter(name==P2_NAME & metric=="Weight") %>% head(1) %>% pull(value)
p2_latest_weight <- df_tidy %>% filter(name==P2_NAME & metric=="Weight") %>% tail(1) %>% pull(value)
p2_wt_lost_as_pct <- -1*100*((p2_latest_weight-p2_starting_weight)/p2_starting_weight)

p1_target_achieved_pct <- (p1_starting_weight-p1_latest_weight)/(p1_starting_weight-P1_TARGET_WEIGHT)*100
p2_target_achieved_pct <- (p2_starting_weight-p2_latest_weight)/(p2_starting_weight-P2_TARGET_WEIGHT)*100
flog.info(glue("p1_starting_weight={p1_starting_weight}, p1_latest_weight={p1_latest_weight},
                p2_starting_weight={p2_starting_weight}, p2_latest_weight={p2_latest_weight},
                p1_target_achieved_pct={p1_target_achieved_pct}, p2_target_achieved_pct={p2_target_achieved_pct}"), name=LOGGER)


# daily weight loss, this is important for a lot of charts and tables
# not the use of group by (name) and then lag. The dataframe is already sorted
# in asc order of time, so if the weight is reducing the daily_wt_loss would be a 
# -ve number, for several charts and tables this is multiplied with -1 so provide
# the absolute loss
df_daily_wt_loss <- df_tidy %>%
  filter(metric == "Weight") %>%
  group_by(name) %>%
  mutate(daily_wt_loss = value - lag(value))

# how many days did it take for each pound to drop? This is found by finding the max date i.e. the last date
# on which each weight (as a whole number, so 230, 229 etc) was seen and then subtracting that date from
# the last date of the previous highest weight. So if 230 was say the 20th pound to drop (if we started from 250 say)
# then the number of days between 231 and 230 becomes the number of days it took to lose the 20th pound.
df_days_to_next_drop <- df_daily_wt_loss %>%
  mutate(value = floor(value)) %>%
  ungroup() %>%
  group_by(name, value) %>%
  summarize(Date=max(Date)) %>%
  arrange(desc(Date)) %>%
  mutate(value_diff=value-lag(value), days=abs(as.numeric(Date-lag(Date)))) %>%
  replace_na(list(value_diff = 0, days = 0)) %>%
  mutate(value=value-min(value)) %>%
  filter(value != 0)

# read the precalculated forecasts and target achievement data 
# this is needed because shinyapps.io does not support Prophet (in the sense there are errors in installing it)
df_forecast_p1 <- read_csv(P1_FORECAST_FPATH) %>%
            select(y, yhat, yhat_lower, yhat_upper, ds)
# convert to xts as that is why dygraph likes
df_forecast_p1 <- xts(df_forecast_p1 %>%
                        select(-ds),
                      order.by = df_forecast_p1$ds)
df_target_achieved_p1 <- read_csv(P1_TARGET_ACHIEVED_FPATH)
df_forecast_p2 <- read_csv(P2_FORECAST_FPATH) %>%
            select(y, yhat, yhat_lower, yhat_upper, ds)
df_forecast_p2 <- xts(df_forecast_p2 %>%
                        select(-ds),
                      order.by = df_forecast_p2$ds)
df_target_achieved_p2 <- read_csv(P2_TARGET_ACHIEVED_FPATH)

flog.info("all done with the initial data read and wrangling", name=LOGGER)
```

Summary
=====================================  

Column {data-width=500}
-----------------------------------------------------------------------
  
### How did we do? Just give me the easy plots.
```{r}
# plot some basic charts, since we have multiple metrics for each date
# so a faceted timeseries would be useful.

#########################################
# NOTE: this applies to all the charts in this dashboard except the timeseries forecasts, so we follow certain
# styling guidelines, these are: 
# 1. Use the 538 theme (yes the one from fivethirtyeight.com)
# 2. Use the tableau color scheme
# 3. Use 20 point font size unless an exception is needed
# 4. Title, subtitle and caption are all mandatory with the caption generally conveying the source of the data.
#########################################

renderPlot({
  df_tidy %>%
  ggplot(aes(x=Date, y=value, col=name)) +
  geom_line() +
  facet_wrap(~metric, scales = "free_y") + 
  theme_fivethirtyeight() +
  xlab("") +
  labs(title=glue(MAIN_PAGE_CHART_TITLE),
       subtitle=glue("Tracking weight and other biometrics, Timespan: {min(df_tidy$Date)} to {max(df_tidy$Date)}"),
       caption=CAPTION) + 
  theme(text = element_text(size=CHART_ELEMENT_TEXT_SIZE_MOBILE), legend.title = element_blank()) + 
  scale_color_tableau()
})
```

Column {data-width=200}
-----------------------------------------------------------------------
  
### How much of our target have accomplished already?

The following gauges show how much of their weight loss targets have `r P1_NAME` and `r P2_NAME` achieved till now.

[]: # (The following code is a clever trick to display two gauges side by side, otherwise they display one below the 
other which wastes precious real estate.)

<div class = "row">
<div class = "col-md-6">
<center>
```{r gauge, fig.width=2, fig.height=2}
rate <- round(p2_target_achieved_pct, 2)
gauge(rate, min = 0, max = 100, symbol = '%', label=P2_NAME,
      gaugeSectors(success = SUCCESS_RANGE, warning = WARNING_RANGE, danger = DANGER_RANGE))
```
</center>
</div>

<div class = "col-md-6">
<center>    
```{r gauge2, fig.width=2, fig.height=2}
rate <- round(p1_target_achieved_pct, 2)
gauge(rate, min = 0, max = 100, symbol = '%', label=P1_NAME,
      gaugeSectors(success = SUCCESS_RANGE, warning = WARNING_RANGE, danger = DANGER_RANGE))
```    
</center>
</div>
</div>


###

```{r}
render_gt ({

  # we want to create a dataframe which contain multiple different metrics which can all
  # be displayed in one single table. We do this by creating a single row dataframe for each
  # metric of interest and then row binding (concatenating) each of these individual dataframes.
    
  # starting weight. Since the weight loss dataframe is sorted
  # in asc order of the date so the first date for each group (name in this case)
  # is automatically the row that contains the starting weight. This way
  # we have a two row dataframe which we can spread to convert to a single row dataframe
  # such that the group i.e. name of person becomes a column name and then we add a new column
  # saying this is the metric that this single row dataframe is tracking. This same strategy
  # of using spread and adding a metric column is followed for all individual dataframes.
  df_starting_wt <- df_wt_loss %>%
    mutate(rn=row_number()) %>%
    filter(rn==1) %>%
    mutate(value=as.character(value)) %>%
    select(name, value) %>%
    spread(name, value) %>%
    mutate(Metric="Starting Weight (lb)") %>%
    select(Metric, everything())
  
  # current weight dataframe  
  df_curr_wt <- df_wt_loss %>%
    mutate(rn=row_number()) %>%
    filter(rn==max(rn)) %>%
    mutate(value=as.character(value)) %>%
    select(name, value) %>%
    spread(name, value) %>%
    mutate(Metric="Current Weight (lb)") %>%
    select(Metric, everything())
  
  # weight loss summary i.e. how much have we lost  
  df_wt_loss_summary <- df_wt_loss %>%
    mutate(rn=row_number()) %>%
    filter(rn==1 | rn==max(rn)) %>%
    mutate(total_wt_loss = as.character(-1*(value-lag(value)))) %>%
    drop_na() %>%
    select(name, total_wt_loss) %>%
    spread(name, total_wt_loss) %>%
    mutate(Metric="Total weight loss (lb)") %>%
    select(Metric, everything())
  
  # total duration  
  df_duration_summary <- df_wt_loss %>%
    mutate(rn=row_number()) %>%
    filter(rn==1 | rn==max(rn)) %>%
    mutate(days_since_start = as.character((Date-lag(Date)))) %>%
    drop_na() %>%
    select(name, days_since_start) %>%
    spread(name, days_since_start) %>%
    mutate(Metric="Days since start") %>%
    select(Metric, everything())
  
  # duration for losing the last 10 pounds
  # this is a tricky one, trying to do everythingin a single dplyr pipeline
  # what we do is as follows: calc the cumulative loss from the latest date
  # and then assign a row number, the rn==1 corresponds to latest date so we
  # want that entry, then we want to find the first entry for which cumul loss >= (say) 10 pounds
  # so we use filter to keep either rn==1 or cumul_loss >= 10, but we only want to keep
  # the first entry where cumul loss beame >=10 so we assign another
  # row number and now we want to the first two rows and then take a diff of the two
  # rows to get number of days. Best to run this outside of shiny to understand.
  df_days_for_last_n_pounds <- df_tidy %>%
    filter(metric=="Weight") %>%
    select(name, Date, value) %>%
    group_by(name) %>%
    arrange(desc(Date), .by_group=TRUE) %>%
    mutate(loss_per_day = value-lag(value, 1)) %>%
    drop_na() %>%
    mutate(cumul_loss = cumsum(loss_per_day)) %>%
    arrange(desc(Date)) %>%
    mutate(rn=row_number()) %>%
    filter(cumul_loss >= N_FOR_LAST_N_POUNDS_OF_INTREST | rn==1) %>%
    mutate(rn2=row_number()) %>%
    filter(rn==1 | rn2==2) %>%
    select(name, Date) %>%
    mutate(days_taken_to_lost_last_n_pounds = as.character(-1*(Date-lag(Date)))) %>%
    drop_na() %>%
    select(name, days_taken_to_lost_last_n_pounds) %>%
    ungroup() %>%
    spread(name, days_taken_to_lost_last_n_pounds) %>%
    mutate(Metric=glue("Days taken to lose last {N_FOR_LAST_N_POUNDS_OF_INTREST} pounds")) %>%
    select(Metric, everything())
  
  # which month did we have the max weight loss (remember weight loss has already been converted to a positive value
  # so loss is +ve and gain is -ve at this point)
  df_best_wt_loss_month <- df_wt_loss %>%
    mutate(m = MONTH_ABB[month(Date)]) %>%
    group_by(name, m) %>%
    summarize(total_wt_loss = sum(loss_per_day)) %>%
    filter(total_wt_loss == max(total_wt_loss, na.rm=TRUE)) %>%
    mutate(best_wt_loss_month = paste0(m, ", ", round(total_wt_loss, 2), " lb")) %>%
    select(name, best_wt_loss_month) %>%
    ungroup() %>%
    spread(name,best_wt_loss_month) %>%
    mutate(Metric="Best Weight loss month") %>%
    select(Metric, everything())
  
  
  # join all the dataframes and display them in the table
  df_imp_metrics <- bind_rows(df_duration_summary,
            df_days_for_last_n_pounds,
            df_starting_wt,
            df_curr_wt,
            df_wt_loss_summary,
            df_best_wt_loss_month)

  gt(df_imp_metrics) %>%
    tab_header(
      title = md("**Important Metrics**"),
      subtitle = md("Key data points that describe the journey")
    )  %>%
    tab_source_note(
      source_note = md(CAPTION)
    ) %>%
    tab_footnote(
      footnote = md(glue("{round(p2_wt_lost_as_pct, 2)}% of the starting body weight.")),
      locations = cells_body(columns = (P2_NAME), rows = 5
    )) %>%
    tab_footnote(
      footnote = md(glue("{round(p1_wt_lost_as_pct, 2)}% of the starting body weight.")),
      locations = cells_body(columns = (P1_NAME), rows = 5
    ))
})
```



Analysis {data-orientation=rows}
=====================================     
   
Row {data-height=1000}
-------------------------------------


### Let us annotate the journey with important events
Some events clearly cause the weight loss to slow down or speed up.
```{r}
renderPlot({
  # We join the important events dataframe with the weight loss dataframe
  # and display the timeseries as a faceted plot
  df_tidy %>%
  filter(metric=="Weight") %>%
  left_join(important_dates ,
            by=c("name","Date")) %>%
  ggplot(aes(x=Date, y=value, col=name)) +
  geom_line() +
  facet_wrap(~name, scales = "free_y", ncol=1) + 
  theme_fivethirtyeight() +
  xlab("") +
  labs(title=glue("The weight loss journey"),
  subtitle=glue("Weight loss & important events, Timespan: {min(df_tidy$Date)} to {max(df_tidy$Date)}"),
  caption=CAPTION) + 
  theme(text = element_text(size=CHART_ELEMENT_TEXT_SIZE_MOBILE), legend.position="none") + 
  theme(axis.title = element_text()) + ylab('Weight (pounds)') +
  scale_color_tableau() + 
  geom_label_repel(aes(label = label),
                   nudge_y = NUDGE_Y,
                   # nudge_x=-3*NUDGE_X,
                   segment.size  = 0.3,
                   arrow = arrow(length = unit(0.05, "npc"), type = "closed", ends = "last"),
                   force = 10,
                   na.rm = TRUE)
})

```

Row {data-height=700}
-------------------------------------
   
### Visualize how each pound was lost
Was it easy to lose the first few pounds, was it much harder to lose the last few pounds? 
```{r}
renderPlot({
  p <- df_days_to_next_drop %>%
    ggplot(aes(x=value, y=days, col=name)) +
    geom_line() +
    geom_point() +
    theme_fivethirtyeight() +
    labs(title=glue("How many days did it take to lose the n'th pound?"),
    subtitle=glue("Timespan: {min(df_tidy$Date)} to {max(df_tidy$Date)}"),
    caption=CAPTION) + 
    theme(axis.title = element_text()) + ylab('Days') + xlab("Pounds lost") +
    theme(text = element_text(size=CHART_ELEMENT_TEXT_SIZE_MOBILE), legend.position="bottom", legend.title = element_blank()) + 
    scale_color_tableau()
  p
})
```   
    
### Were some days of the week better than others in terms of weight loss?
Is there a "curse of the weekend?" i.e. no weight loss or even a weight gain on Monday or Tuesday following indiscretions over the weekend?
```{r}
renderPlot({
  
  df_daily_wt_loss %>%
  mutate(weekday = wday(Date, label = TRUE)) %>%
  group_by(name, weekday) %>%
  summarize(wt_loss = median(daily_wt_loss, na.rm=TRUE)) %>%
  ggplot(aes(x=weekday, y=-1*wt_loss, fill=name, col=name)) +
  geom_bar(position="dodge", stat="identity") +
  # facet_wrap(~name) + 
  theme_fivethirtyeight() +
  xlab("") + 
  labs(title=glue("Median weight loss by day of week"),
       subtitle=glue("Timespan: {min(df_tidy$Date)} to {max(df_tidy$Date)}"),
       caption=CAPTION) + 
  theme(axis.title = element_text()) + ylab('Weight loss (lb)') +
  theme(text = element_text(size=CHART_ELEMENT_TEXT_SIZE_MOBILE), legend.position = "bottom") + 
  scale_fill_tableau() +
  scale_color_tableau()
})
```

Forecasts
=====================================

Column {data-width=500}
-----------------------------------------------------------------------
  
### Forecasts for `r P1_NAME`

```{r}
renderDygraph ({
  # render the timeseries chart for the forecast using the precalculated data
  dygraph(df_forecast_p1,
          glue("Weight loss forecast for {P1_NAME}")) %>%
    dySeries("y", label = "Actual") %>%
    dySeries(c("yhat_lower", "yhat", "yhat_upper"), label = "Predicted") %>%
    dyLimit(df_target_achieved_p1$target[1],
            glue("Target: {df_target_achieved_p1$target[1]} lb, Expected date: {df_target_achieved_p1$date[1]}")) %>%
    dyAnnotation(df_target_achieved_p1$date[1],
             text = substr(P1_NAME, 1, 1),
             tooltip = glue("Target: {df_target_achieved_p1$target[1]} lb, Expected date: {df_target_achieved_p1$date[1]}"))
})
```

### Forecasts for `r P2_NAME`

```{r}
renderDygraph ({
  # render the timeseries chart for the forecast using the precalculated data
  dygraph(df_forecast_p2,
          glue("Weight loss forecast for {P2_NAME}")) %>%
    dySeries("y", label = "Actual") %>%
    dySeries(c("yhat_lower", "yhat", "yhat_upper"), label = "Predicted") %>%
    dyLimit(df_target_achieved_p2$target[1],
            glue("Target: {df_target_achieved_p2$target[1]} lb, Expected date: {df_target_achieved_p2$date[1]}")) %>%
    dyAnnotation(df_target_achieved_p2$date[1],
                 text = substr(P2_NAME, 1, 1),
                 tooltip = glue("Target: {df_target_achieved_p2$target[1]} lb, Expected date: {df_target_achieved_p2$date[1]}"))
 
})
```

About
=====================================

```{r}
includeMarkdown(ABOUT_FILE)
```
