---
title: "Biometric Tracker"
output:
  flexdashboard::flex_dashboard: null
  orientation: columns
  vertical_layout: fill
runtime: shiny
resource_files:
- raw_data/*
- data/*
---
  
```{r setup, include=FALSE}
library(flexdashboard)
```

Summary
=====================================  

Column {data-width=500}
-----------------------------------------------------------------------
  
### How did we do? Just give me the easy plots.
```{r}
```

Column {data-width=200}
-----------------------------------------------------------------------
  
### How much of our target have accomplished already?



###

```{r}
```



Analysis {data-orientation=rows}
=====================================     
   
Row {data-height=500}
-------------------------------------


### Let us annotate the journey with important events
Some events clearly cause the weight loss to slow down or speed up.
```{r}

```

Row {data-height=500}
-------------------------------------
   
### Visualize how each pound was lost
Was it easy to lose the first few pounds, was it much harder to lose the last few pounds? 
```{r}
```   
    
### Were some days of the week better than others in terms of weight loss?
Is there a "curse of the weekend?" i.e. no weight loss or even a weight gain on Monday or Tuesday following indiscretions over the weekend?
```{r}
```

Forecasts
=====================================

Inputs {.sidebar data-width=300}
-----------------------------------------------------------------------

Forecast parameters for 

```{r}
```

<br>

Forecast parameters for 

```{r}
```


Column {data-width=500}
-----------------------------------------------------------------------
  
### Forecasts for 

```{r}
```

### Forecasts for 

```{r}
```

About
=====================================

```{r}
```


